### 2019-05-06 papers/haskell19/1-intro.tex
Pass over abstract and Section 1

% --------------------------------------
% Intro
%
Random property-based testing is a quite powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16}.
%
In the Haskell comunity, QuickCheck is the predominant tool of this such.
%
The developers specify
%
\begin{inparaenum}
\item the testing properties their systems must fullfil, and
\item random data generators for the data types involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random seeds using these generators, and uses them to
evaluate the system's properties in search of possible counterexamples, which
often indicate bugs in our system.

% --------------------------------------
% Motivating Example
%
Although QuickCheck provides default random generators for the common base types
like |Int| or |String|, it requires implementing random generators for any
user-derined data type we want to randomly generate.
%
% This section introduces the common workflow for generating random values using
% automatically derived QuickCheck generators, along with its common limitations.
%
To illustrate this, let us consider the following data type definition encoding
basic HTML pages:

\begin{code}
data Html
  =  Text  String
  |  Sing  String
  |  Tag   String Html
  |  Join  Html Html
\end{code}

This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| to simply concatenate two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, and in the scope of random generation in general, we will refer
to any non-recursive data constructor as \emph{terminal}, and
\emph{non-terminal} in any other case.

Then, using this representation, the example page\linebreak
%
\texttt{<html><b>hello<br>world!</b></html>}
%
can be encoded simply as:
%
\begin{code}
Tag "html" $ Tag "b" $
  Text "hello" iJoin Sing "br" iJoin Text "world!"
\end{code} %$

% --------------------------------------
% Random Generators
%
Then, in order to generate random values of type |Html|, we need to provide a
random data generator for it.
%
For this purpose, QuickCheck defines the |Arbitrary| type-class for default
defining random generators as:

@@

\begin{code}
instance Arbitrary Html where
  arbitrary = sized gen
    where  gen 0 = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary) ]
           gen d = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary)
             ,  (4,  Tag     <$> arbitrary  <*> gen (d-1))
             ,  (3,  Join    <$> gen (d-1)  <*> gen (d-1)) ]
\end{code} %$
%
At the top level, this definition parametrizes the generation process using
QuickCheck's |sized| combinator, which lets us build our generator using an
auxliary local function |gen :: Int -> Gen Html|.
%
The |Int| passed to |gen| is known as the \emph{generation size}, and is
threaded seamlessly by QuickCheck on each call to arbitrary.
%
Then, we use this parameter to limit the maximum amount of recursive calls that
our generator can perform, and thus the maximum depth of the generated values.
%
If the generation size is positive (case |gen d|), our random generator picks a
random constructor from our |Html| data type with a given generation frequency
(using QuickCheck's |frequency| combinator), and proceeds to fill its fields
using randomly generated sub-terms---here using Haskell's applicative notation
\cite{Mcbride2008} and the default |Arbitrary| instance for |String|s.
%
For the case of the recursive sub-terms, this generator simply calls itself
recursively with a smaller generation size (|gen (d-1)|).
%
This process repeats until we reach the base case (|gen 0|) on each recursive
sub-term.
%
At this point, our random generator is limited to only pick among the |Html|
terminal constructors, hence terminating the generation process.

% --------------------------------------
% Automatic derivation tools
%
As one can observe, the previous definition is quite mechanical, and depends
only on the generation frequencies we choose for each constructor.
%
Fortunately, there exists a variety of meta-programming tools for obtaining
random generators automatically based on their types.
%
Tools like \emph{MegaDeTH} or \emph{DRAGEN} will synthesize random generators
using the same procedure as the one shown above.
%
\emph{MegaDeTH} simply picks among data constructors uniformingly, whereas
\emph{DRAGEN} analyzes the type definitions and tunes the generation frequencies
based on a simulation-based optimization approach, in order to obtain random
distributions of values based on the users' desires.

% --------------------------------------
% Limitation
%
Such tools provide a useful mechanism for automating the boring and error-prone
task of writing random generators by hand.
%
However, this simple derivation procedure, which only considers the structural
information encoded in data type definitions, and generates one single
constructor at a time; is often too generic to synthesize useful generators in
some practical scenarios.
%
For instance, because of the types we use for deriving our random generators are
too general, leaving a large room for ill-formed values, e.g., invalid HTML
pages; or alternatively, because our properties need to be tested against very
specific patterns of input cases in order to trigger bugs.

This problem, however, can be alleviated by considering additional sources of
structural information from our codebase other than just data constructors.
%
For instance, having a set of high-level combinators, defining an abstract
interface of our data types, e.g., |hr :: Html| or |bold :: Html -> Html|; it is
possible to include them into the generation process, thus obtaining more
complex and structed values.
%
In the same way, having functions defined using pattern matching over the data
types of our interest, e.g. |f (Join (Text t) (Tag s x)) = dots|, it is also
possible to include this rich structure which exposes a relationship among
different data constructors into our random generators.
%
This idea is addressed in depth in the next section, while the rest of this work
focuses on implementing it in a compositional way.

%
% Contributions
%
The main contribution of this paper are:
%
\begin{itemize}
\item We identify some important limitations of performing random testing with
  automatically derived generators using state-of-the-art tools (Section
  \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different source constructions using Data Types \`a la Carte (Section
  \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We present a very simple yet powerful domain specific language for
  describing extensible generators solely based on the types used to represent
  the desired shape of our random data (Section \ref{sec:typelevel}).
\item We provide a Template Haskell tool for automatically deriving all the
  required machinery presented throughout this paper, and evaluate its
  generation performance using three real-world case studies (Section
  \ref{sec:casestudies}).
\end{itemize}

>>>>

% --------------------------------------
% Intro
%
Random property-based testing is a powerful technique for finding bugs
\cite{ArtsHNS15, HughesPAN16, HughesNSA16}.
%
In the Haskell community, QuickCheck is the predominant tool of this such.
%
The developers specify
%
\begin{inparaenum}[(i)]
\item the testing properties their systems must fulfill, and
\item random data generators (or generators for short) for the data types
  involved at their properties.
\end{inparaenum}
%
Then, QuickCheck generates random seeds using these generators, which are used
to evaluate the system's properties in search of possible counterexamples, which
often indicate the presence of bugs.

% --------------------------------------
% Motivating Example
%
Although QuickCheck provides default generators for the common base types, like
|Int| or |String|, it requires implementing generators for any user-defined data
type we want to generate.
%
% This section introduces the common workflow for generating random values using
% automatically derived QuickCheck generators, along with its common limitations.
%
To illustrate this, let us consider the following data type definition encoding
basic HTML pages:

\begin{code}
data Html  =  Text  String
           |  Sing  String
           |  Tag   String Html
           |  Join  Html Html
\end{code}
%
This type allows us to build HTML pages via four possible data constructors:
|Text| is used for plain text values; |Sing| and |Tag| represent singular and
paired HTML tags, respectively; and |Join| simply concatenates two HTML pages
one after another.
%
Note that the constructors |Tag| and |Join| are recursive, as they have at least
one field of their own type.
%
In this paper, we will refer to any non-recursive data constructor as
\emph{terminal}, and \emph{non-terminal} in any other case.

Using this representation, the example page \linebreak
%
\texttt{<html><b>hello<br>world!</b></html>}
%
is encoded as:
%
\begin{code}
Tag "html" $ Tag "b" $
  Text "hello" iJoin Sing "br" iJoin Text "world!"
\end{code} %$

% --------------------------------------
% Random Generators
%
As mentioned before, in order to generate random |Html| values, we need to
provide a random generator for it.
%
For this purpose, QuickCheck defines the |Arbitrary| type-class for defining
default generators as:

@@

\begin{code}
instance Arbitrary Html where
  arbitrary = sized gen
    where  gen 0 = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary) ]
           gen d = frequency
             [  (2,  Text    <$> arbitrary)
             ,  (1,  Sing    <$> arbitrary)
             ,  (4,  Tag     <$> arbitrary  <*> gen (d-1))
             ,  (3,  Join    <$> gen (d-1)  <*> gen (d-1)) ]
\end{code} %$
%
At the top level, this definition parametrizes the generation process using
QuickCheck's |sized| combinator, which lets us build our generator via an
auxiliary local function |gen :: Int -> Gen Html|.
%
The |Int| passed to |gen| is known as the \emph{generation size}, and is
threaded seamlessly by QuickCheck on each call to arbitrary.
%
Then, we use this parameter to limit the maximum amount of recursive calls that
our generator can perform, and thus the maximum depth of the generated values.
%
If the generation size is positive (case |gen d|), our random generator picks a
random |Html| constructor with a given generation frequency (using QuickCheck's
|frequency| combinator), and proceeds to fill its fields using randomly
generated sub-terms---here using Haskell's applicative notation
\cite{Mcbride2008} and the default |Arbitrary| instance for |String|s.
%
For the case of the recursive sub-terms, this generator simply calls |gen|
recursively with a smaller generation size (|gen (d-1)|).
%
This process repeats until we reach the base case (|gen 0|) on each recursive
sub-term.
%
At this point, the generator is limited to pick only among terminal |Html|
constructors, hence ending the generation process.

% --------------------------------------
% Automatic derivation tools
%
As one can observe, the previous definition is quite mechanical, and depends
only on the generation frequencies we choose for each constructor.
%
Fortunately, there exists a variety of meta-programming tools for obtaining
random generators automatically, based on the information encoded on the
definition of the types we want to generate.
%
Tools like \emph{MegaDeTH} or \emph{DRAGEN} will synthesize random generators
using roughly the same procedure as the one shown above.
%
On one hand, \emph{MegaDeTH} derives generators which simply pick among data
constructors with uniform probabilities, whereas \emph{DRAGEN} analyzes type
definitions, tuning the generation frequencies based on a predictive mechanism.
%
This allows the users to specify the desired distribution of values of the
derived generators.

% --------------------------------------
% Limitation
%
These tools provide a useful mechanism for automating the repetitive and
error-prone task of writing random generators by hand.
%
However, this simple derivation procedure, which
%
\begin{inparaenum}[(i)]
\item only considers the structural information encoded in type definitions, and
\item generates one single constructor at a time;
\end{inparaenum}
%
is often too generic to synthesize useful generators in practical scenarios.
%
For instance, because of the types we use for deriving our generators are too
general, leaving a large room for ill-formed values, e.g., invalid HTML pages;
or alternatively, because our properties need to be tested against very specific
input patterns in order to trigger bugs.

This problem, however, can be alleviated by considering additional sources of
structural information from our codebase other than just type definitions
\cite{Mista2019GeneratingRS}.
%
For instance, having a set of high-level combinators, defining an abstract
interface of our data types, e.g., |hr :: Html| or |bold :: Html -> Html|; it is
possible to include them into the generation process, thus obtaining more
complex and structured values.
%
In the same way, having functions defined using pattern matching over the data
types of our interest, e.g. |f (Join (Text t) (Tag s x)) = dots|, it is also
possible to include this rich structure which exposes a relationship among
different data constructors into our random generators.
%
This idea is addressed in depth in the next section, while the rest of this work
focuses on implementing it in a compositional way.

%
% Contributions
%
The main contribution of this paper are:
%
% \vspace{-10pt}
%
\begin{itemize}
\item We identify some important limitations of performing random testing with
  automatically derived generators using state-of-the-art tools (Section
  \ref{sec:sources}).
\item We present an extensible mechanism for representing random values built
  upon different source constructions using Data Types \`a la Carte (Section
  \ref{sec:representation}).
\item We develop a modular generation schema, exploiting the composable nature
  of our representations, which are extended to encode information relevant to
  the generation process at the type level (Section \ref{sec:generators}).
\item We propose a simple type-level idiom for describing extensible generators,
  based on the types used to represent the desired shape of our random data
  (Section \ref{sec:typelevel}).\looseness=-1
\item We provide a Template Haskell tool for automatically deriving all the
  required machinery presented throughout this paper, and evaluate its
  generation performance using three real-world case studies (Section
  \ref{sec:casestudies}).
\end{itemize}

### 2019-05-06 papers/haskell19/abstract.tex
Pass over abstract and Section 1

In this work, we develop an extensible framework for deriving compositional
generators, which can be easily combined in diferent ways in order to fit the
developers' demands, using a simple type-level domain-specific language.
%
Our framework is based on familiar ideas from the \`a la Carte technique for
writing composable interpreters in Haskell, and extended with the required
machinery for it to work in the scope of random generation of values.
%
Additonally, we provide an implementation of our ideas, and evaluate its
performance using real-world examples.

>>>>

In this work, we develop an extensible framework for deriving compositional
generators, which can be easily combined in different ways in order to fit the
developers' demands, using a simple type-level description language.
%
Our framework relies on familiar ideas from the \`a la Carte technique for
writing composable interpreters in Haskell, extended with the machinery required
in the scope of random generation.
%
We provide an implementation of our ideas, and evaluate its performance using
real-world examples.
